Visualising metagenome differences

### 2023-11-21 analysis/metagenome.Rmd
Add the unclassified MAGs too

```{r setup, message=FALSE}
library(tidyverse)
library(UpSetR)
library(microshades)
library(ggtext)
library(patchwork)
library(scales)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

@@

@@

```{r}
amph_pos_tax <- clean_tax |> filter(Sample_species %in% c('Amphibolis antarctica', 'Posidonia australis'))
```

@@

```{r fig.height=6}
plot_orders_gtdb <- function(amph_pos_tax, group_remainder = TRUE) {
  temp <- amph_pos_tax |> 
    group_by(sample, Tissue, Sample_species) |> 
    mutate(Sample_species = trunc_str(Sample_species)) |> 
    mutate(Sample_species = paste0('*', Sample_species, '*')) |> 
    mutate(order = str_replace_all(order, 'o__','')) |> 
    count(order, .drop = FALSE) 
  if(group_remainder) {
    temp <- temp |> 
    mutate(counted_orders = case_when(n == 1 ~ 'Remainder',
                                      TRUE ~ order))
  } else {
      temp <- temp |> 
        mutate(counted_orders = order)
  }
  temp <- temp |> 
    select(-order) |> 
    group_by(sample, Tissue, Sample_species, n, counted_orders) |> 
    summarise(n = sum(n)) |> 
    filter(counted_orders != '' )

  uniques <- c(setdiff(unique(temp$counted_orders), c('Remainder')), 'Remainder')
  temp |> 
    mutate(counted_orders = factor(counted_orders, levels=uniques)) |> 
    mutate(sample = str_remove(sample, "[A-Z]+")) |> 
    mutate(sample = str_replace(sample, '31', '3')) |> 
    ggplot(aes(x = factor(sample, levels=rev(unique(sample))), y = n, fill=counted_orders)) +
    geom_col() + 
    theme_minimal() + 
    facet_wrap(~Tissue+Sample_species, scales = 'free_y', ncol=2) + 
    coord_flip() + 
    ylab('Number of contigs') + xlab('Replicate') +
    labs(fill = 'Order') +
    scale_y_continuous(breaks=pretty_breaks())  +
    # move legend position to bottom
    theme(legend.position = "bottom",
          strip.text.x = element_markdown()) +
    scale_fill_manual(values = colors)
}
plot_orders_gtdb(amph_pos_tax)
```
Let's also make an Upset plot of that

```{r}
tmp <- amph_pos_tax |> 
  group_by(sample) |> 
  mutate(order = str_replace_all(order, 'o__','')) |> 
  count(order) |> 
  filter(order != '' ) |>
  filter ( n > 1) |> 
  summarise(named_vec = list(order)) %>%
  deframe()
tmp
```
```{r}
upset(fromList(tmp))
```

@@

clean_tax_cat <- clean_tax_cat |> left_join(metadata, by = c('sample'='Sample'))
```

@@

```{r fig.height=6}
plot_orders_cat <-
  function(amph_pos_tax_cat, group_remainder = TRUE) {
    temp <- amph_pos_tax_cat |>
      separate(Order, into = c('Order', 'Order_score')) |>
      rename(order = Order) |>
      group_by(sample, Tissue, Sample_species) |>
      mutate(Sample_species = trunc_str(Sample_species)) |>
      mutate(Sample_species = paste0('*', Sample_species, '*')) |>
      count(order)
    if (group_remainder) {
      temp <- temp |>
        mutate(counted_orders = case_when(n == 1 ~ 'Remainder',
                                          TRUE ~ order))
    } else {
      temp <- temp |>
        mutate(counted_orders = order)
    }
    temp <- temp |>
      select(-order) |>
      group_by(sample, Tissue, Sample_species, n, counted_orders) |>
      summarise(n = sum(n)) |>
      filter(counted_orders != '')

    uniques <-
      c(setdiff(unique(temp$counted_orders), c('Remainder')), 'Remainder')
    temp |>
      mutate(counted_orders = factor(counted_orders, levels = uniques)) |>
      mutate(sample = str_remove(sample, "[A-Z]+")) |>
      mutate(sample = str_replace(sample, '31', '3')) |>
      ggplot(aes(
        x = factor(sample, levels = rev(unique(sample))),
        y = n,
        fill = counted_orders
      )) +
      geom_col() +
      theme_minimal() +
      facet_wrap( ~ Sample_species + Tissue, ncol = 2, scales = 'free_y') +
      coord_flip() +
      ylab('Number of contigs') + xlab('Replicate') +
      labs(fill = 'Order') +
      scale_y_continuous(breaks = pretty_breaks())  +
      # move legend position to bottom
      theme(legend.position = "bottom",
            strip.text.x = element_markdown()) +
      scale_fill_manual(values = colors)
  }
plot_orders_cat(amph_pos_tax_cat)
```

>>>>

```{r setup, message=FALSE}
library(tidyverse)
library(microshades)
library(ggtext)
library(scales)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

@@

Need to put some extra work in; MAGs with no classification are not in the output file.

```{r}
all_mag_counts <- read_tsv('./data/metagenome/MAG_counts.tsv') |> mutate(Sample = str_replace(Sample, '_L2', '')) |> 
  rename(TotalMAGs = MAGs)
```

@@

```{r}
amph_pos_tax <- clean_tax |> filter(Sample_species %in% c('Amphibolis antarctica', 'Posidonia australis')) |> 
  left_join(all_mag_counts, by = c('sample'='Sample'))
```

@@

```{r fig.height=6}
plot_orders_gtdb <- function(amph_pos_tax, group_remainder = TRUE) {
  temp <- amph_pos_tax |> 
    group_by(sample, Tissue, Sample_species,TotalMAGs) |> 
    mutate(Sample_species = trunc_str(Sample_species)) |> 
    mutate(Sample_species = paste0('*', Sample_species, '*')) |> 
    mutate(order = str_replace_all(order, 'o__','')) |> 
    count(order, .drop = FALSE) 
  if(group_remainder) {
    temp <- temp |> 
    mutate(counted_orders = case_when(n == 1 ~ 'Remainder',
                                      TRUE ~ order))
  } else {
      temp <- temp |> 
        mutate(counted_orders = order)
  }
  temp2 <- temp |> 
    select(-order) |> 
    group_by(sample, Tissue, Sample_species, n, counted_orders) |> 
    summarise(n = sum(n)) |> 
    filter(counted_orders != '' )
  # we need to count the unclassified MAGs that are not in the output
  group_presents <- temp2 |> group_by(sample) |> summarise(classifiedMAGs = sum(n))
  missings <- group_presents |> 
    left_join(temp |> 
                select(sample, TotalMAGs)) |> 
    unique() |> 
    mutate(unclassifiedMAGs = TotalMAGs - classifiedMAGs) |> 
    select(-c(classifiedMAGs,TotalMAGs)) |>
    rename(Unclassified = unclassifiedMAGs) |>
    pivot_longer(-c(sample, Tissue, Sample_species)) |> 
    rename(counted_orders=name, n=value)

  temp2 <-rbind(temp2, missings)

  uniques <- c(setdiff(unique(temp2$counted_orders), c('Unclassified', 'Remainder')), 'Remainder', 'Unclassified')
  temp2 |> 
    mutate(counted_orders = factor(counted_orders, levels=uniques)) |> 
    mutate(sample = str_remove(sample, "[A-Z]+")) |> 
    mutate(sample = str_replace(sample, '31', '3')) |> 
    ggplot(aes(x = factor(sample, levels=rev(unique(sample))), y = n, fill=counted_orders)) +
    geom_col() + 
    theme_minimal() + 
    facet_wrap(~Tissue+Sample_species, scales = 'free_y', ncol=2) + 
    coord_flip() + 
    ylab('Number of contigs') + xlab('Replicate') +
    labs(fill = 'Order') +
    scale_y_continuous(breaks=pretty_breaks())  +
    # move legend position to bottom
    theme(legend.position = "bottom",
          strip.text.x = element_markdown()) +
    scale_fill_manual(values = colors)
}
plot_orders_gtdb(amph_pos_tax)
```

@@

clean_tax_cat <- clean_tax_cat |> left_join(metadata, by = c('sample'='Sample')) |> left_join(all_mag_counts, by =c('sample'='Sample'))
```

@@

```{r fig.height=6}
plot_orders_cat <-
  function(amph_pos_tax_cat, group_remainder = TRUE) {
    temp <- amph_pos_tax_cat |>
      separate(Order, into = c('Order', 'Order_score')) |>
      rename(order = Order) |>
      group_by(sample, Tissue, Sample_species, TotalMAGs) |>
      mutate(Sample_species = trunc_str(Sample_species)) |>
      mutate(Sample_species = paste0('*', Sample_species, '*')) |>
      count(order)
    if (group_remainder) {
      temp <- temp |>
        mutate(counted_orders = case_when(n == 1 ~ 'Remainder',
                                          TRUE ~ order))
    } else {
      temp <- temp |>
        mutate(counted_orders = order)
    }
    temp2 <- temp |>
      select(-order) |>
      group_by(sample, Tissue, Sample_species, n, counted_orders,TotalMAGs) |>
      summarise(n = sum(n)) |>
      filter(counted_orders != '')
      # we need to count the unclassified MAGs that are not in the output
    group_presents <- temp2 |> group_by(sample) |> summarise(classifiedMAGs = sum(n))
    missings <- group_presents |> 
      left_join(temp2) |> 
      select(classifiedMAGs, Tissue, Sample_species, sample, TotalMAGs)|> 
      unique() |> 
      mutate(unclassifiedMAGs = TotalMAGs - classifiedMAGs) |> 
      select(-c(classifiedMAGs,TotalMAGs)) |>
      rename(Unclassified = unclassifiedMAGs) |>
      pivot_longer(-c(sample, Tissue, Sample_species)) |> 
      rename(counted_orders=name, n=value)

    temp2 <-rbind(temp2, missings)

   uniques <- c(setdiff(unique(temp2$counted_orders), c('Unclassified', 'Remainder')), 'Remainder', 'Unclassified')

    temp2 |>
      mutate(counted_orders = factor(counted_orders, levels = uniques)) |>
      mutate(sample = str_remove(sample, "[A-Z]+")) |>
      mutate(sample = str_replace(sample, '31', '3')) |>
      ggplot(aes(
        x = factor(sample, levels = rev(unique(sample))),
        y = n,
        fill = counted_orders
      )) +
      geom_col() +
      theme_minimal() +
      facet_wrap( ~ Sample_species + Tissue, ncol = 2, scales = 'free_y') +
      coord_flip() +
      ylab('Number of contigs') + xlab('Replicate') +
      labs(fill = 'Order') +
      scale_y_continuous(breaks = pretty_breaks())  +
      # move legend position to bottom
      theme(legend.position = "bottom",
            strip.text.x = element_markdown()) +
      scale_fill_manual(values = colors)
  }
plot_orders_cat(amph_pos_tax_cat)
```
